Agreement Attraction in Turkish: Effects of Nominal and Verbal Plural Morphemes

### 2019-08-06 Paper/paper_draft.rmd
Started working on the paper. [PL]

      @TannerBulkes:2015
      @Staub:2010
      @DankEtAl:2014 **
      @PearlmutterGarnseyBock:1999
      @FooteBock:2012
      @DillonEtAl:2013

@@

It has been observed that speakers generally fail to accurately comprehend agreement between elements in a sentence. One such comprehension error that is made has been deemed agreement attraction. Speakers may erroneously find a sentence in which the "attractor" agrees with the verb instead of the "head" of the subject. Data from typologically different languages and explanations from different frameworks show that I am far from understanding by what the attraction is modulated. Some theories refer this phenomenon as a faulty representation of the number on the head of the subject phase whereas other theories argue that the attraction is due to an error in the access to the number of the subject. In this paper, I report two experiments exploring another possible explanation for agreement attraction: an extreme shallow processing. The first experiment is a replication of @Lago2018 where I changed the head of the subject in order to differentiate between the accusative case *-I*. In the second experiment, I have used a relative clause which only consists of a plural marked verb as a distractor.  my results show that while people do indeed care more than just "matching" the forms of plural verbal. either comprehenders have some sort of restriction to not use cues from verbs upon accessing the number of the plural morpheme or there is a syntactic reason for the number feature of the plural morpheme to not "percolate."

# Literature

## Theories of attraction

Agreement attractions are defined as instances of incorrect agreements of an element with a distractor that is not its grammatical controller. As in \autoref{E1}, the target of the agreement is the verb *are*, the attracting distractor is *bottles*, and the feature erroneously assessed is the number feature of the grammatical controller *label*.

\begin{exe}
\ex[*]{The label on the bottles \textit{are} wrong.\footnotemark}
\label{E1}
\end{exe}

\footnotetext{The convention of using $\ast$ is to show the ungrammaticality of the sentencee.}

@@

In their study, @Lago2018 investigates the comprehension of subject-verb agreement  Turkish monolinguals. They use speeded acceptability judgments for the effects of number attraction in Turkish. Their sentences make us of genitive-possessive constructions in the subject position, where the genitive is the attractor and the possessive is the head noun. They manipulate the grammaticality of the sentence by changing the plural morphology of the verb, and they also manipulate the plurality of the attractor noun. In grammatical conditions, the subject and the verb both bear singular morphology with no overt morpheme. Moreover, in the ungrammatical conditions, the verb bears the overt *-lAr* morpheme whereas the subject is still singular as exemplified below.

@@

\ex \textsc{Ungrammatical, SG attractor} \label{lago4}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singer's backup vocalist jumped on the stage non-stop.

\ex \textsc{Grammatical, PL attractor} \label{lago2}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{Ungrammatical, PL attractor} \label{lago3}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{Grammatical, SG attractor} \label{lago1}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singer's backup vocalist jumped on the stage non-stop.

@@

\begin{exe}
\ex
  \begin{xlist}
  \ex \underline{Ungrammatical, PL attractor} \label{exp3}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}--\textsc{pl}\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{Grammatical, PL attractor} \label{exp2}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{Ungrammatical, SG attractor} \label{exp4}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-\textsc{pl}\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

  \ex \underline{Grammatical, SG attractor} \label{exp1}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

@@

One hundred and seven Turkish speakers were recruited from Bogazici University in İstanbul. I did not collect participants' knowledge of other languages; however, I verified that Turkish is their native language and that they predominantly use it in their daily lives. In the experiments, participants were asked to judge the acceptability of experimental and filler sentences in Turkish. All of the sentences were presented one word at a time in the center of the screen for 500 ms per word unlike @Lago2018 and @Wagers2009, in which the duration was 300 ms per word. The experiment was run on a web-based platform titled the Ibex Farm, and all of the documentation can be found on  my Github page:\url{https://github.com/utkuturk/AgreementAttraction_Experiment}.

Before the experiment, participants were instructed to give accurate and quick answers based on their own intuitions, and they were also notified about the time limit for answering. At the start of the experiment, they were given 4 practice items with feedback. Each trial after the practice items started with a blank screen, followed by a flash of single words in the middle of the screen. After the sentence, they were asked a grammaticality judgment questions (Was the sentence you read was natural to you?) where they need to press P for *no* and press Q for *yes*. In the experiment, every participant saw 40 experimental item that is Latin-squared and 40 fillers.

Every experimental item from both experiments (Experiment 1 and @Lago2018) is included in the analysis. Participants that could not exceed the threshold of 0.25 difference in *yes* answer between the condition (c) and (d), namely Ungrammatical, SG attractor and Grammatical, SG attractor, were excluded. I performed an analysis by fitting linear mixed-effect models to  my data, controlling for vowel ending, plural attractor, grammaticality and their interactions. Models for responses were fit using Bernoulli distribution with probit link function and 95\% confidence intervals. Models for response times were fit using Lognormal distribution.

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup

>>>>

      * @TannerBulkes:2015
      * @Staub:2010
      * @DankEtAl:2014 **
      * @PearlmutterGarnseyBock:1999
      * @FooteBock:2012
      * @DillonEtAl:2013
      - Shravan et al.'s new paper on Armenian

@@

It has been observed that speakers often fail to accurately process grammatical dependencies between the parts of a sentence. 
For example, \ldots illusory npi licensing, agreement attraction \ldots
One such comprehension error that is made has been dubbed agreement attraction (cite): Speakers may erroneously find sentences acceptable, in which the the verb erroneously agrees with an unreated noun phrase (*the attractor*) instead of the subject (*the agreement controller*).

Some theories (cite) explain this phenomenon by means of a faulty representation of the number on the head of the agreement controller whereas other theories assume that the attraction is due to an error in the access to the number of the subject.

In this paper, we report two experiments exploring another possible explanation for agreement attraction: shallow processing. The first experiment is a replication of @Lago2018 in which a potentially confounding variable is controlled. In the second experiment, we explore whether verbs with plural agreement can serve as as potential attractors.  The results show that 
<!--
pvl: no idea what the text below means
while people do indeed care more than just "matching" the forms of plural verbal. either comprehenders have some sort of restriction to not use cues from verbs upon accessing the number of the plural morpheme or there is a syntactic reason for the number feature of the plural morpheme to not "percolate." 
-->

# Theories of attraction

Agreement attraction are defined as instances of incorrect agreements of an element with a distractor that is not its grammatical controller. As in \autoref{E1}, the target of the agreement is the verb *are*, the attracting distractor is *bottles*, and the feature erroneously assessed is the number feature of the grammatical controller *label*.

\begin{exe}
\ex[*]{The key to the cabinets \textit{are} \ldots }
\label{E1}
\end{exe}

@@

In their study, @Lago2018 investigate the comprehension of subject-verb agreement  Turkish monolinguals. They use speeded acceptability judgments for the effects of number attraction in Turkish. Their sentences make us of genitive-possessive constructions in the subject position, where the genitive is the attractor and the possessive is the head noun. They manipulate the grammaticality of the sentence by changing the plural morphology of the verb, and they also manipulate the plurality of the attractor noun. In grammatical conditions, the subject and the verb both bear singular morphology with no overt morpheme. Moreover, in the ungrammatical conditions, the verb bears the overt *-lAr* morpheme whereas the subject is still singular as exemplified below.

@@

\ex \textsc{ungrammatical, sg attractor} \label{lago4}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singer's backup vocalist jumped on the stage non-stop.

\ex \textsc{grammatical, pl attractor} \label{lago2}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{ungrammatical, sg attractor} \label{lago3}
\gll \c{S}ark{\i}c{\i}-lar-{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}-lar.\\
singer-\textsc{pl}-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-\textsc{3Pl}\\
\glt The singers' backup vocalist jumped on the stage non-stop.

\ex \textsc{grammatical, pl attractor} \label{lago1}
\gll \c{S}ark{\i}c{\i}-n{\i}n vokalist-i sahne-de s\"{u}rekli z{\i}pla-d{\i}\\
singer-\textsc{gen} vocalist-\textsc{poss} stage-\textsc{loc} non-stop jump-\textsc{pst}-$\varnothing$\\
\glt The singer's backup vocalist jumped on the stage non-stop.

@@

\begin{exe}
\ex
  \begin{xlist}
  \ex \underline{ungrammatical, pl attractor} \label{exp3}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}--\textsc{pl}\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{grammatical, pl attractor} \label{exp2}
      \gll Komedyen-ler-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{pl}-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedians' assistant felt chilly.

  \ex \underline{ungrammatical, sg attractor} \label{exp4}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}-ler.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-\textsc{pl}\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

  \ex \underline{grammatical, sg attractor} \label{exp1}
      \gll Komedyen-in yard{\i}mc{\i}-s{\i} poyraz-dan dolayi \"{u}\c{s}\"{u}-d\"{u}.\\
  comedian-\textsc{gen} helper-\textsc{poss} northeaster-\textsc{abl} because.of feel.chilly-\textsc{pst}-$\varnothing$\\
      \glt Because of the northeaster, comedian's assistant felt chilly.

@@

One hundred and seven Turkish speakers were recruited from Bogazici University in İstanbul. We verified that Turkish was their native language and that they predominantly used it in their daily lives. 
In the experiments, participants were asked to judge the acceptability of Turkish sentences. All of the sentences were presented one word at a time in the center of the screen for 500 ms per word.
<!-- unlike @Lago2018 and @Wagers2009, in which the duration was 300 ms per word. -->
The experiment was run on *Ibex Farm*, a web-based experiment platform (\url{http://spellout.net/ibexfarm}). All experimental stimuli, data, and documentation can be found on \url{https://github.com/utkuturk/AgreementAttraction_Experiment}.

Participants were instructed to quickly clasify sentences as acceptable or unacceptable based on their intuition. They were further informed about the $XXXX ms$ time limit for answering. At the start of the experiment, they were given 4 practice trials, and received feedback on their response. Each trial after the practice items started with a blank screen, followed by a word-by-word RSVP presentaion of the sentence in the center of the screen. After the sentence presentation, they were asked for a grammaticality judgment 
(Was the sentence you read was natural to you?) 
<!-- Is this really how it was phrased? -->
responded Q for *yes*, and P for *no*. 
The 40 experimental items were automatically divided into four lists according to a Latin-square design, and were intermixed with filler sentences, such that each participant saw 40 experimental items and 40 fillers in random order over the course of the experiment.

### Data Analysis
<!-- to-do: find out if the items from experiment 1 and Lago's paper were minimal pairs. Should
at least some of them have the same item id? -->

In order to find out whether the noun class, and thus the presence of an ambiguity between genitive and accusative affected the magnitude of the agreement attraction effect, we analyzed the data from our experiment
together with the data from @Lago2018.

Participants for whom the difference in the percentage of `yes' responses between grammatical and ungrammatical conditions without attraction (conditions c and d) was below $0.25$ were deemed to not have performed the task as instructed, and their data was excluded from further analysis.

We used `R` `r citep(citation())` and the *tidyverse* R packages `r citep(citation("tidyverse"))` for data processing and plotting, and the R packages *brms* `r citep(citation("brms"))` and *rstan* `r citep(citation("rstan"))` to fit Bayesian hierarchical models [e.g.,@GelmanHill:2007, @McElreath:2016, @BDA3].

A Bayesian hierarchical generalized linear model was fitted to the participant's responses

I performed an analysis by fitting linear mixed-effect models to  my data, controlling for vowel ending, plural attractor, grammaticality and their interactions. Models for responses were fit using Bernoulli distribution with probit link function and 95\% confidence intervals. Models for response times were fit using Lognormal distribution.

@@
